---
title: "Project Time series"
output: html_document
date: "2026-04-15"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is the code document used to do the time series analysis in the
context of the project from the Time Series course on IST-UL. The
authors are João Cordeiro (ist1107480) and Susana Ferreira (ist59482).

# Assignment 1

## Southern Oscillation Index (SOI) since 2010 monthly data

```{r}
url <- "https://raw.githubusercontent.com/JoaoCordeiro2004/Time-Series-Project/refs/heads/main/Time%20Series%20SOI%202010-2025.txt"

SOI <- read.csv(url, header = FALSE)
```

The first transformations we do with the raw data is to convert the data
to an actual data format and create column names for the data frame.

```{r}
colnames(SOI) <- c("date", "soi")
SOI$date <- as.Date(paste0(SOI$date, "01"), format = "%Y%m%d")
```

Afterwards, we create the time series for the data and plot it.

```{r}
soi_ts <- ts(SOI$soi,
             start = c(2010, 1),
             frequency = 12)
```

```{r}
plot(SOI$date, SOI$soi,
     type = "l",
     main = "Southern Oscillation Index (SOI)",
     xlab = "Time",
     ylab = "SOI",
     col = "blue")
```

Next, we handle the missing values in the Time Series

```{r}
library(imputeTS)

ggplot_na_distribution(soi_ts)
```

As we can see by the graph there are no missing values for this
particular Time Series, therefore there was no need to impute the gaps.

With the time series being completely clean, we want to fit a SARIMA
model to be able to do a forecast. For a first analysis, we do a STL
decomposition of the original time series.

```{r}
stl_soi <- stl(soi_ts, s.window = "per")

plot(stl_soi)
```

There is no clear trend pattern, but there seems to be a seasonal
component. Since this data originates from climate measures, we can
expect a seasonal lag of 12 months. (????)

To determine if the original time series needs any transformation, it is
necessary to inspect if it is homogeneous. For that, we analyse the ACF
and PACF, as well as run some statistical tests.

```{r}
library(tseries)
acf(soi_ts, lag.max = 100)
pacf(soi_ts, lag.max = 100)

adf.test(soi_ts)
pp.test(soi_ts)
kpss.test(soi_ts)
```

Looking at the ACF function we can see a slow decay and a peak in the
PACF function. Also, when the ADF test is considered we can conclude
that we don't have enough information to reject the null hypothesis
(not-stationary), but the PP and KPSS tests tell us the contrary.
Therefore the provided data can't give a straightforward conclusion
about the stationarity of time series.

With this in mind, we consider some models where the data is transformed
with difference and some without said difference and analyse them. We
consider **d = 1** and **d = 0,** with **s=12**. Also, we consider a
model with seasonal difference **D = 1** and **s = 12**.

#### *Model with d=1, s=12* 

#### *Q = 0 , P = 2 ,  p = 0 , q = 1 and D = 0.*

```{r}
library(tseries)
library(forecast)

soi_ts_1 <- diff(soi_ts, 1)

auto.arima(soi_ts_1)

plot.ts(soi_ts_1)

acf(soi_ts_1, lag = 50)
pacf(soi_ts_1, lag = 50)

adf.test(soi_ts_1)
pp.test(soi_ts_1)
```

#### *Model with d=0, s=12* 

#### *Q = 1 , P = 2 , p = 2 , q = 1 and D = 0.*

```{r}
library(tseries)
library(forecast)

auto.arima(soi_ts)

plot.ts(soi_ts)

acf(soi_ts_1, lag = 50)
pacf(soi_ts_1, lag = 50)
```

**Model with D = 1, s = 12.**

**Q = 0 , P = 2 , q = 0 , p = 2 and d = 0**

**Q = 0 , P = 2 , q = 1 , p = 1 and d = 0**

```{r}

soi_ts_2 = diff(soi_ts, 12)

adf.test(soi_ts_2)
pp.test(soi_ts_2)

auto.arima(soi_ts_2)

plot.ts(soi_ts_2)

acf(soi_ts_2, lag = 50)
pacf(soi_ts_2, lag = 50)

```

**ISTO A SEGUIR ESTÁ ERRADO!!!!**

We proceed to decompose it into its Seasonal, Trend and Residual
components. As per usual, we assume the cyclical component to be zero.

```{r}

```

### Analysis of the Remainder Component

Having performed the STL decomposition, we can extract the remainder
component of the SOI time series. The goal is to fit a SARIMA model to
it.

```{r}
soi_rem <- stl_soi$time.series[,3]

plot(soi_rem)
```

Before proceeding, we need to check if this time series is stationary.
Otherwise, we couldn't proceed with the modelling.

```{r}
library(tseries)
adf.test(soi_rem)
kpss.test(soi_rem)
```

By Augmented Dickey-Fuller test (null hypothesis states that it is not
stationary) and the KPSS test (null hypothesis states that the TS is
stationary) we work with the hypothesis that the remainder component is
stationary.

A SEGUIR DAR PLOT DAS ACF E PACF E TIRAR CONCLUSÕES. EXPLICAR NO
RELATÓRIO TUDO DEPOIS!!!!

------------------------------------------------------------------------

```{r}
diff1 <- diff(soi_ts, differences = 1)

```

```{r}
diff_seasonal <- diff(diff1, lag = 12)
plot(diff_seasonal)
adf.test(diff_seasonal)
```

```{r}
acf(diff_seasonal)
pacf(diff_seasonal)
```

lag 1 -\> positive to autocorrelation ACF spikes-\> suggest MA(q) or
MA(Q) PACF spikes -\> suggest AR(p) or AR(P)

```{r}
library(forecast)

model_sarima <- auto.arima(soi_ts)
summary(model_sarima)
```

Diagnostic checking: No autocorrelation in residuals (ACF) -\> white
noise Ljung-Box test p-value \> 0.05 data: Residuals from ARIMA(1,0,1)
with zero mean Q\* = 16.228, df = 22, p-value = 0.8044 Model df: 2.
Total lags used: 24

The Ljung-Box test is a statistical diagnostic tool used in time series
analysis to determine if residuals from a fitted model (ARIMA) are
independently distributed, meaning they exhibit no significant
autocorrelation, or if they behave as white noise. It tests the overall
randomness of the data across multiple lags rather than a single lag
{\displaystyle H\_{0}}: The data is not correlated (i.e. the
correlations in the population from which the sample is taken are 0, so
that any observed correlations in the data result from randomness of the
sampling process). {\displaystyle H\_{a}}: The data exhibit serial
correlation.

```{r}
checkresiduals(model_sarima)
```

After deciding on appropriate models, forecast the data into the future
up to 5 time periods ahead and calculate 95% prediction intervals for
each of the 5 forecasts.

```{r}
forecast_5 <- forecast(model_sarima, h = 5, level = 95)
plot(forecast_5,
     main = "SOI Forecast (5 steps ahead)",
     ylab = "SOI",
     xlab = "Time")
```

# Assigment 2

```{r}
library(readxl)

EDP <- read_excel(
  "C:/Users/X1605/OneDrive - IPLeiria/Ano_letivo_2025_26/master_ist/Time_Series_Analysis/project/EDP.xls",
  skip = 3
)

GALP <- read_excel(
  "C:/Users/X1605/OneDrive - IPLeiria/Ano_letivo_2025_26/master_ist/Time_Series_Analysis/project/GALP ENERGIA-NOM.xls",
  skip = 3
)

MENGIL<-read_excel(
  "C:/Users/X1605/OneDrive - IPLeiria/Ano_letivo_2025_26/master_ist/Time_Series_Analysis/project/MOTA ENGIL.xls",
  skip = 3
)

MENGIL<-read_excel(
  "C:/Users/X1605/OneDrive - IPLeiria/Ano_letivo_2025_26/master_ist/Time_Series_Analysis/project/MOTA ENGIL.xls",
  skip = 3
)

Navigator<-read_excel(
  "C:/Users/X1605/OneDrive - IPLeiria/Ano_letivo_2025_26/master_ist/Time_Series_Analysis/project/THE NAVIGATOR COMP.xls",
  skip = 3
)

```

## Log-returns associated to the daily closing values for the 4 TS

```{r}
library(dplyr)

EDP <- EDP %>%
  arrange(Date) %>% 
  mutate(log_return = log(Close / lag(Close)))
```
